Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2013-01-24 documents/mixtures/response_to_referee.txt

This was a very useful referee report and we have revised the paper
accordingly.  I think the referee comments have improved the paper
substantially.

David W. Hogg (on behalf of the authors)

-----

We very much appreciate the thorough and sensible referee comments;
they have led to substantial improvements in the paper.  In what
follows, the referee comments are indented and our responses are not
indented.

@@

   1. Does the article contain significant new results and/or analysis
   and reflect sufficiently high scientific standards to warrant its
   publication in PASP?

   This paper gives a particularly lucid, concise and clear
   description of a relatively old idea. As such, it represents a
   significant contribution to the literature as a pedagogical
   reference. It should be published but see 4. below.

We agree with the referee that the idea of using Gaussians to fit
galaxies is old but we don't agree with any implication that the
method or applied math results shown are old.  All references we can
find in the literature show *how* to do mixture-of-Gaussian fitting in
real imaging, but don't give mixture-of-Gaussian approximations to
known profiles.  This paper, in that sense, is more of an "applied
math meets astronomy" paper than the previous work.  We will return to
this below, where we *do* make changes to the text in response to this
general concern of the referee.

DSTN NOTES:.. math convenience... pre-compiling in Photo... adaptive
sampling challenges... Bundy and Marshall uses... programming
complexity... These points need to go in the paper, not just this
response!

@@

   Page 6: Why is "badness" computed in a one-dimensional numerical integral? For
   the sake of computational speed? What happens when dealing with heavily blended
   objects? Where is the center for the 1D integral then? Please explain. A general
   explanation of how object deblending is performed with MoGs would be welcome.

   Page 7, paragraph 2: The statement "These -- M^lux = 6 and M^luv = 8 -- are good
    compromises between mixture complexity and quality of fit" warrants more
   explanation. Everytime I read about using MoGs for galaxy fitting, I have the same
   reaction: it needs a large number of highly redundant to model something a single
   exponential profile. Fans of the eigenimage decompositions are probably appalled!
    Granted the mathematical simplicity of MoGs can make up for this by enabling
   faster fitting times, but it is definitely an un-appealing aspect of the method.
   It would be helpful to comment on this to counter a reaction that I think may be
   shared by others. More importantly, I do not quite understand how you decided on
    M^lux = 6 and M^luv = 8. If I look at Table 2, the badness for lux has clearly
   converged and there is no gain in going higher than M=6. I cannot say the same
   thing for the luv MoG. Badness goes down by a factor of 16 going from M=6 to M=8
   and then goes down by a factor of 5 going from M=8 to M=10. How can you claim
   that your badness has "converged" for M=8 then? A quantitative threshold and its
   rationale are clearly needed here to justify the "good compromises" here. Please
   be more quantitative or clarify if I misunderstood something.

   Equation 19: I did not quite understand your distinction between the \Delta X
   vector here and the x vector in equation 2. Please clarify.

In other news, we also made the following changes to the text, in
response to other comments and issues:

- we did foo

- we did bar

### 2013-02-10 documents/mixtures/response_to_referee.txt

We very much appreciate the thorough and sensible referee comments;
they have led to substantial improvements in the paper.  In what
follows, the referee comments are indented and our responses are not
indented.

   1. Does the article contain significant new results and/or analysis
   and reflect sufficiently high scientific standards to warrant its
   publication in PASP?

   This paper gives a particularly lucid, concise and clear
   description of a relatively old idea. As such, it represents a
   significant contribution to the literature as a pedagogical
   reference. It should be published but see 4. below.

We agree with the referee that the idea of using Gaussians to fit
galaxies is old but we don't agree with any implication that the
method or applied math results shown are old.  All references we can
find in the literature show *how* to do mixture-of-Gaussian fitting in
real imaging, but don't give mixture-of-Gaussian approximations to
known profiles.  This paper, in that sense, is more of an "applied
math meets astronomy" paper than the previous work.  We will return to
this below, but we *did* make changes to the text in response to this
general concern of the referee: We added sentences in the introductory
paragraphs and at the end of the paper to emphasize the difficulty of
rendering the profiles without these approximations.

To be more specific: we added some sentences to emphasize that the
currently used methods for rendering profiles are effectively mixtures
of delta functions (the profile is sampled at a finite set of
locations) and rendering requires as many PSF renderings as there are
sample points.  This is a huge computation.

We also added an introductory paragraph that points out that the code
complexity required to do rendering correctly is large and this leads
to engineering overheads in testing and correctness that go way beyond
the simple numerical speed losses.

[should we also add a late paragraph re-iterating and showing a comparison?]

@@

   3. Could the order or presentation or English be improved?  Editing
   of English and typographic errors will be corrected by the Journal
   copyediting staff and you do not need to specify these in your
   report.  But if there are more general problems with presentation
   or English please cite this in your report.

   Contractions (don't,  doesn't) should not be used.

Replaced all contractions.

   On page 8n, paragraph 3: "If your PSF is not in MoG form ..." should
   read "If the PSF is not in MoG form ..."

@@

Fixed to "If the PSF is not represented in MoG form ...".

   4. Do you have any comments or criticisms that may be helpful to
   the author(s) to improve or correct the paper? In cases that may be
   ambiguous, please specify whether you consider the suggested
   changes to be mandatory for publication or advisory.

   GENERAL COMMENT: Although I do think that this "Note" would stand
   on its own as a separate publication, I would nonetheless strongly
   suggest that the authors consider merging it into the forthcoming
   Lang et al. paper on The Tractor. This Note appears to be an "ad"
   for The Tractor. In the words of the authors, the main point of the
   Note is to emphasize the performance gain that the MoG approach
   would bring to the analysis of truly large galaxy datasets.
   However, the current draft does not show any analysis of real data
   to support this performance claim.  Other methods have been
   proposed in the past with similar promises of significant
   performance improvements (e.g., fast nonlinear regression via
   eigenimages, Brigham et al.). They looked promising on paper but
   were found to exhibit unexpected biases when applied to real
   data. I would hate to see this happen to MoGs. This is why it would
   be really nice to have the beautiful description of MoGs presented
   in this Note side-by-side with some tests on real data in the same
   paper.

As noted above, we have added sentences in one paragraph and one new
paragraph that makes it clearer what the performance gains are.

[ should we also get quantitative at the end? ]

@@

   SPECIFIC COMMENTS:

   Page 2, paragraph 3: "There is no theoretical argument and only
   weak observational arguments that the rotation-supported parts of
   galaxies are always exponential and that the kinematically hot
   components are always more de Vaucouleurs-like".  Although your
   warning about over-interpretating exp and dev components using
   bulges and disks is well taken (and others have given the same
   warning I might add), I think your statement above is too
   strong. On the theoretical side, some works (e.g., Brooks et
   al. 2011, ApJ, 728, 51; Agertz et al. 2011, MNRAS, 410, 1391) have
   performed photometric decompositions of their model galaxies to see
   how photometric components mapped onto kinematical ones. The
   agreement is good and does suggest that it is often a reasonable
   mapping. It is certainly not a "radical" assumption to make. Also,
   while it is true that some dispersion-supported galaxies such as
   dwarf ellipticals have exponential profiles, they certainly do not
   make up the bulk of the galaxy population in large
   surveys. Finally, I note the recent ATLAS3D paper by Krajnovic et
   al. 2012, arXiv:1210.816 in which it was found that "early-type"
   galaxies with significant rotational support also exhibited
   significant exponential ("disk") components.

Point well taken!  We softened this paragraph so that it no longer
makes any claim on this point; we now only attempt to induce some
feelings about conservativeness.

### 2013-02-10 documents/mixtures/mixture_models.tex
preample trivialities
notes to self
one more paragraph of response to referee

% This file is part of The Tractor.
% Copyright 2012 David W. Hogg (NYU) and Dustin Lang (Princeton, CMU).
% All rights reserved.

\documentclass[12pt,pdftex,preprint]{aastex}
% \documentclass[12pt,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs}

@@

Whenever an investigator is fitting PSF-convolved exponential or
de~Vaucouleurs profiles, the models presented here will improve code
performance.  That does not mean that doing such fitting is a Good
Idea.  These profiles are effective models for galaxies at low
signal-to-noise (as they are used in the \project{SDSS}) and they are
useful templates for performing consistent photometry across
morphologically diverse objects (as we use them in \citealt{bundy}).
More radically, these profiles are sometimes used to distinguish disk
and bulge light (as in \citealt{bulgedisk} and references cited
therein); these uses are prone to over-interpretation: There is no
theoretical argument and only weak observational arguments that the
rotation-supported parts of galaxies are always exponential and that
the kinematically hot components are always more de-Vaucouleurs-like.

The performance advantages we will obtain from using MoG
approximations are not simply that convolution itself is trivial.  The
standard unconvolved galaxy models, especially the de~Vaucouleurs and
S\'ersic models (\citealt{dev}, \citealt{ser}), are very ill-behaved
near the galaxy center.  Rendering these profiles precisely near the
center can be very challenging numerically.  In addition, most of the
rendering time is spent at very small radii, where PSF convolution is
going to erase all structure anyway.  That is, a MoG description of
the de~Vaucouleurs profile saves time both in rendering \emph{and} in
PSF convolution; it produces profiles that are approximate but very
high performance when all real uses of the profiles are PSF-convolved,
as they usually are.

>>>>

% This file is part of The Tractor.
% Copyright 2012, 2013 David W. Hogg (NYU) and Dustin Lang (Princeton, CMU).
% All rights reserved.

% to-do items
% -----------
% - respond to referee
% - resubmit

\documentclass[12pt,pdftex,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs}

@@

Whenever an investigator is fitting PSF-convolved exponential or
de~Vaucouleurs profiles, the models presented here will improve code
performance.  That does not mean that doing such fitting is a Good
Idea.  These profiles are effective models for galaxies at low
signal-to-noise (as they are used in the \project{SDSS}) and they are
useful templates for performing consistent photometry across
morphologically diverse objects (as we use them in \citealt{bundy}).
More radically, these profiles are sometimes used to distinguish disk
and bulge light (as in \citealt{bulgedisk} and references cited
therein); these uses are prone to over-interpretation: There certainly
could be rotation-supported parts of galaxies that are not exponential
and kinematically hot components that are not de-Vaucouleurs-like.

The performance advantages we will obtain from using MoG
approximations are not simply that it makes the convolution trivial.
The standard unconvolved galaxy models, especially the de~Vaucouleurs
and S\'ersic models (\citealt{dev}, \citealt{ser}), are very
ill-behaved near the galaxy center.  Rendering these profiles
precisely near the center can be very challenging numerically.
Computer codes that do this rendering and PSF convolution well use
enormous numbers of samples; mathematically this sampling
representation of the models is a mixture of delta functions.  In
order for these representations of the profiles to be precise, large
numbers of sample points are required.  The convolution and rendering
step involves as many PSF evaluations as there are samples---which can
number in the thousands.  In addition, most of the rendering time is
spent at very small radii---most of the samples in any integration are
in or near the central pixel---where PSF convolution is going to erase
all structure anyway.  That is, a MoG description of the
de~Vaucouleurs profile saves time both in rendering \emph{and} in PSF
convolution; it produces code that is very high performance when all
real uses of the profiles are PSF-convolved, as they usually are.

Not only does a sampling representation of the profiles require many
more compute operations per render than a MoG representation, it also
requires more code complexity.  For the sampling to be both accurate
and efficient, it has to be strongly adaptive (with much higher
density in the center).  This adds complexity and choices into the
code, leading to engineering and software maintenance and testing
overheads that cost not just compute time but human time.  One
response to this is to build large number of pre-computed profile PSF
convolutions and interpolate among them.  This is the strategy taken
in the \project{SDSS} pipelines (\citealt{lupton}), but that requires
even \emph{more} human overhead, because not only does the sampling
and convolution have to be done correctly, so does the numerical
interpolation.
